Machine learning guided batched design\\ of a bacterial Ribosome Binding Site

### 2021-12-18 paper/main.tex
cheng's changes in intro

Optimisation of gene expression levels is an essential part of the organism design process.
Fine control of this process can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
Unfortunately, design of specific genetic parts can still be challenging due to lack of reliable design methods.
To address this problem, we have created a machine learning guided Design-Build-Test-Learn cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed \hl{using relatively small, high-quality data sets}.
We used Gaussian Process Regression for the Learn part of cycle and the Upper Confidence Bound multi-armed Bandit algorithm for the Design of genetic variants to be tested in vivo.
We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for a more reliable data generation.
Notably, by testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding our benchmark RBS by up to 34\%.
Overall, our results show that machine learning is a powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.\\

@@

One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
\hl{Part design, understood as modifying the sequence of a part for it to meet specific requirements, is a central problem of synthetic biology. This is} because, it is the genetic parts that are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms. 
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
This cycle involves designing the DNA sequence in computer-aided design (CAD) software and then physically testing it in a laboratory.
Additionally, computer modelling can be used to predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.
\\

\hl{For instance}, according to  \textcite{Reeve2014} there are three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models range from relatively good ($R^2 >0.8$) to low ($R^2 <0.2$) depending on the data set \cite{Reis2020}.
This may be due to number of reasons: 
i) they rely on calculations of free energies which can be difficult to estimate with high precision, 
ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and 
iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which can potentially increase prediction error \cite{Goss1998}. 
There is also evidence that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}. 
This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.
\hl{This suggests that there is a number of interactions determining the mRNA-ribosome binding which can be hard to capture in a single, universal model}.\\

The volume of experimental data in synthetic biology is increasing exponentially \cite{Freemont2019}. 
Recent work has combined these data sets with machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}. 
For example, Jervis \emph{et al.} used support vector machines and neural networks to optimise production of monoterpenoids in \emph{Escherichia coli} \cite{Jervis2019}.
Similarly, Costello \emph{et al.} have used a number of machine learning approaches to analyse time-series multiomics data to predict metabolic pathway behaviour \cite{Costello2018}. 
Deep learning techniques have also been successfully used to analyse large synthetic biology data sets \cite{Alipanahi2015,Angermueller2016,Hollerer2020}.
\hl{However, such big data sets are still generated relatively rarely and there is a need to develop methods operating on smaller data sets.}\\

\hl{In genetic design practice, the designers often have to fine tune characteristics of the used parts to meet the specific metrics of the designed strains.
For instance, if the goal is to produce a protein at a high yield, the designer might want to increase the TIR of the RBS responsible for translation of the said protein.
To do that, one could use the existing RBS predictors, but they can be unreliable in the investigated genetic context.
On the other hand, the designer could use machine learning methods to make the predictions, but these usually require large data sets.
To address this problem, we have created a machine learning guided Design-Build-Test-Learn cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed using even relatively small, but high-quality data sets.}
Importantly, this work focuses on RBS part design and characteristic prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
As the RBS is one of the key genetic elements controlling protein expression and has a relatively short sequence, it is a suitable target for establishing a workflow that could be potentially translated to more complicated systems.\\

Our overall experimental goal was to maximise the TIR by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
\maciej{I've decided not to put the "other than maximisation" sentence here, I think it's appropriate to have it in discussion.}
We demonstrate how machine learning (ML) algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level \hl{in the specific genetic context (i.e. with the specific bases upstream and downstream of the investigated RBS sequences)}. 
Two types of ML algorithms are applied in the DBTL cycle. 
For the Learn part, the goal is design a predictor to predict the protein expression level by learning from the logged data. 
We used a Bayesian, non-parametric regression algorithm, called Gaussian Process Regression \cite{Rasmussen2004} in our pipeline. 
\hl{GPR, as a non-parametric Bayesian regression algorithm, assumes that the reward function is sampled from a Gaussian Process and has been shown to  perform well with a small amount of data in biological prediction tasks} \cite{Rasmussen2004, desautels2014parallelizing}. 
For the Design part, the goal is to design a policy to recommend RBS sequences to query sequentially in batches so that we can identify the optimal RBS sequences within a given budget. 
A variant of the Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} was applied in our pipeline to guide our recommendations. 
The two types of algorithms cooperate with each other and provides powerful tools for DBTL cycle in genetic parts design \cite{srinivas2012information, desautels2014parallelizing}.
\hl{This way, we were able to rationally increase production of our target protein compared to the very strong benchmark RBS by up to 35\%}. \\

@@

\hl{The exact position of the RBS in the sequence upstream of the CDS can be hard to pinpoint, however most previous studies put it in the 20 bp directly preceding the coding sequence.} 
In our investigation, we are using an RBS which is known to have a very high TIR when expressing GFP and is present in the pBb series of plasmids \cite{Lee2011}.
This template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACAT, where the highlighted nucleotides constitute the core of the RBS.
Since this is the sequence against which new RBS sequences will be benchmarked,
we will refer to this sequence as the \textit{benchmark sequence} hereafter.
Additionally, we have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Figure \ref{fig:core_vs_noncore}).
This hypothesis has been built based on reported biases towards certain bases present in the core of the RBS but absent outside of it.
For example, according to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
Similarly, outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overall TIR of a given RBS. 
Focusing on the 6 to 8 bp core sequence is a common RBS design approach \cite{Jeschek2016}.
Hence, in our design, we focus on modification of the core at nucleotide positions -8 to -13 (relative to the start codon of the GFP; this is where the consensus Shine-Dalgarno AGGAGG sequence is usually found in wild type \textit{E. coli}) of the RBS and we keep other positions the same as the benchmark sequence, i.e. TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T). 
The total experimental (variant) space is then $4^6$ = 4096.\\

@@

Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group shows the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups, suggesting that the algorithms have a better understanding of the experimental space given more data.
The decreased result in the 0.9th quantile compared to the maximum value for the Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown in Supplementary \ref{subsec: Exploration-exploitation visualisation} and Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
\hl{The implication of the fact that an RBS sequence chosen at random will have low TIR
(as shown by UNI and PPM), is that most of the exploration will result in low TIR.
Our results confirm that RBSs with high TIRs tend to come from exploitation of the design space,
whereas the exploration points give relatively low TIRs.
Note that the exploration is necessary to expand our knowledge to the unknown parts of the design space and in effect allow us to exploit it better.}\\

@@

\hl{To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design part, we have selected a regression algorithm, Gaussian Process Regression (GPR),  which can predict not only the means, but also uncertainty estimations for our data points. }
Figure \ref{fig: Scatterplot} \hl{shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round. As expected, the predictions in Round 0 were poor due to the use of approximate data. 
The predictions improved for the subsequent rounds, 
the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.}
% Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.
\\

\hl{The regression task in the DBTL cycle is more challenging than the large-scale data based regression tasks.
In the early iterations, we have a limited number of data points and relatively high variability due to the measurement noise. 
However, since the predictions are only used for recommendations in the Design part, instead of precise predictions of the mean for each RBS, we only need to provide a valid ranking for both the predicted mean TIR and uncertainties, and thus the ranking of the UCB scores.
This would ensure that for each round we would be testing the required designs, even if we are not exactly accurate in terms of their numerical characteristics.
Because of that, we provide the evaluation of the Learn part in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearman correlation coefficient}
\cite{Schober2018,  Kang2019}.\\

Additionally, the 
% obtained values of the coefficient of determination and Spearman correlation coefficient 
performance of Learn part 
is also influenced by the exploration-exploitation balance in the Design part.
In each round, we select some data points for exploration of the areas in which we have few or none tested data points.
For those exploration points, since the predictor never learns their label distributions and patterns, there is no chance for the predictor to provide accurate predictions of TIR values,
% which means that when tested, they have a high chance of having a experimentally-obtained mean different from what was predicted, 
thus hurting the prediction performance in the short term.
However, this is useful information for future predictions as it allows us to understand the whole underlying space instead of focusing on local sub-optimal data points.
\hl{In other words, we are intentionally sacrificing the accuracy of our predictions in each round to improve them in the future rounds}. 
The effect of exploration of the space is the ability to find high TIR RBSs even with relatively low prediction performance.\\

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_0.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_1.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_2.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_3.pdf}
    \end{subfigure}
    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation)}. The scatter plots A-D show the performance of our prediction algorithm calculated after each round.
    Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
    The Spearman correlation coefficient (with corresponding p-value; calculated using test data only) are provided for each plot.
    X-axis and y-axis are the true, tested TIR and the predicted TIR respectively.
    In D), we train our predictor on randomly chosen 80\% of all available data, and test on the remaining 20\% data. 
    The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
    }
    \label{fig: Scatterplot}
\end{figure}

@@

\hl{Finally, while our strong sequences showed some affinity towards the anti-sense sequence of the ribosome known to bind to RBS, they did not show any obvious secondary structures that could explain their TIRs} (see Figure \ref{fig:structures}).
This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion, based on the previously reported literature \cite{Saito2020,EspahBorujeni2016}, that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently \hl{optimise a small genetic part, in this case maximising TIR of a bacterial RBS}. 
We have used Gaussian Process regression to predict the TIR function to be optimised and an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested.
To represent RBSs and capture the similarities between them, we chose to use the Weighted Degree Kernel with Shift method, which fits well into our prediction method.
In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow and testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs \hl{in this context} by up to 34\%.
Furthermore, we have generated an extensive library of diverse RBSs that can be used \hl{as a basis} in future studies.\\

In our work we have focused on maximising the TIR of the RBS, by gradually moving our emphasis from exploration to exploitation as we progressed through the design rounds.
\hl{As long as maximisation could be the appropriate function for optimising RBSs or other small parts}, it may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.
This change could be achieved within the Bayesian optimisation framework by changing the objective of the algorithm to maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objective method, for example \mbox{\cite{shu2020new}}.\\
% \hl{
% While we focus our optimisation framework on the RBS optimisation task, our framework has good potential to be generalised to multi-gene pathway design, as reviewed by \mbox{\textcite{LAWSON2021}} and proposed by \mbox{\textcite{hamedirad2019towards}}. For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process Regression (GPR) prediction model could be updated to a deep GPR approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity \mbox{\cite{shahriari2015taking}}.  
% }

@@

There are still open questions that need to be addressed for applying machine learning in synthetic biology.
Firstly, we would like to understand how we can extract more biologically-important information from the decisions made by our algorithms.
We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the results obtained.
Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty measurements? 
Thirdly, the generalisability of the method is unknown.
We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
\hl{Similarly, the re-usability of the obtained data set is currently unknown. 
The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures} \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} \hl{respectively) studies would indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.}
Finally, the optimal exploration-exploitation balance between rounds and samples is still to be determined.
In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.\\

>>>>

Optimisation of gene expression levels is an essential part of the organism design process.
Fine control of this process can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
Unfortunately, design of specific genetic parts can still be challenging due to lack of reliable design methods.
To address this problem, we have created a machine learning guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs to show how small genetic parts can be reliably designed using relatively small, high-quality data sets.
We used Gaussian Process Regression for the Learn phase of cycle and the Upper Confidence Bound multi-armed bandit algorithm for the Design of genetic variants to be tested in vivo.
We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for reliable data generation.
Notably, by Testing a total of 450 RBS variants in four DBTL cycles, we experimentally validated RBSs with high translation initiation rates equalling or exceeding our benchmark RBS by up to 34\%.
Overall, our results show that machine learning is a powerful tool for designing RBSs, and they pave the way towards more complicated genetic devices.\\

@@

One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
Part design, understood as modifying the sequence of a part for it to meet specific requirements, is a central challenge in synthetic biology. Genetic parts are the units which are ultimately combined into more complicated genetic circuits that produce desired functions in the target organisms. 
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism is continually improved through an iterative process.
This cycle involves designing the DNA sequence in computer-aided design (CAD) software and then physically testing it in a laboratory.
Additionally, computer modelling can be used to predict characteristics of a genetic part \cite{Yeoh2019,Nielsen2016}.
Most of these computer models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically-obtained values describing a relevant design property, like translation initiation rate (TIR) in the case of ribosome binding sites (RBSs) \cite{Xia1998,Chen2013,Reeve2014}.
However, de novo design of small genetic elements is still challenging due to unknown relationships between their sequence and performance.
\\

In the Design phase of the DBTL cycle, designers often have to fine tune characteristics of parts to give the resulting strains their desired properties. The ability to predict a characteristic of a genetic part (for example using a machine
learning method) only solves half of the problem. The designer still needs to choose from the large number of 
possible variants to Build.
For instance, if the goal is to produce a protein at a high yield, the designer might increase the TIR of the RBS responsible for translation of that protein.
Hence the Design phase of TIR has two steps: first to predict the TIR value for each potential RBS sequence,
and second to select a small set of RBS sequences based on their predicted TIR values.\\

To achieve the first step, one could use the existing RBS predictors. \textcite{Reeve2014} surveys three main RBS calculators, all predicting the TIR based on the thermodynamic properties of the RBS and the ribosome \cite{Seo2013,Na2010,Salis2009}. 
Reported predictions from all of these models range from relatively good ($R^2 >0.8$) to low ($R^2 <0.2$) depending on the data set \cite{Reis2020}.
This may be due to number of reasons: 
i) they rely on calculations of free energies which can be difficult to estimate with high precision, 
ii) in general, one of the best ways to improve the models' accuracy is by increasing the number of phenomena taken into account, which can lead paradoxically to decreased model accuracy due to accumulation of errors \cite{EspahBorujeni2016}, and 
iii) by using deterministic coefficients to calculate energies, one disregards the often stochastic nature of processes in cells which can potentially increase prediction error \cite{Goss1998}. 
There is also evidence that binding energy calculations may be poor predictors of RBS strength \cite{Saito2020,Sherer1980}. 
This is reinforced by studies suggesting that RNA secondary structure is potentially a more important feature in TIR determination \cite{DESMIT1994,EspahBorujeni2016}.
This suggests that multiple interactions determine the mRNA-ribosome binding, and predicting TIR from the genomic sequence is still challenging.\\

Recent work has explored the use of machine learning algorithms to generate new models and predictors for use in synthetic biology, vastly improving the DBTL cycle's performance \cite{Camacho2018,Radivojevic2020,LAWSON2021}.
This leverages the exponential increase in experimental data from synthetic biology~\cite{Freemont2019}
to improve predictors in the Learn phase of the DBTL cycle.
For example, \textcite{Jervis2019} used support vector machines and neural networks to optimise production of monoterpenoids in \emph{Escherichia coli}.
Similarly, \textcite{Costello2018} have used a number of machine learning approaches to analyse time-series multi-omics data to predict metabolic pathway behaviour. 
Deep learning techniques have also been successfully used to analyse large synthetic biology data sets \cite{Alipanahi2015,Angermueller2016,Hollerer2020}.
However large data sets may not be available for a genetic part of interest, for instance the RBS. Hence there is a need to develop methods for training predictors on smaller data sets in the Learn phase of DBTL. The trained predictors
can then be used in the first step of the Design phase to predict characteristics of interest for a potential variant.\\

In this paper, we propose a machine learning guided Design-Build-Test-Learn cycle 
for the experimental design of bacterial RBSs. We show how small genetic parts can be reliably designed using even relatively small, but high-quality data sets.
In particular, we develop a method for the Learn phase, and use the learned predictor for the first step of the Design phase.
Furthermore we develop a method for the second step of the Design phase,
that recommends batches of genetic designs that optimize the characteristic of interest.
This work focuses on RBS part design and TIR prediction, rather than looking at its wider genetic context and impact on the general performance of the cell.
As the RBS is one of the key genetic elements controlling protein expression in bacteria,
it is a suitable target for establishing a workflow that could be potentially translated to more complicated systems.\\

Our overall experimental goal is to maximise the TIR by building and testing batches of RBS sequences while minimising the number of DBTL cycle iterations.
%\maciej{I've decided not to put the "other than maximisation" sentence here, I think it's appropriate to have it in discussion.}
We demonstrate how machine learning (ML) algorithms can be incorporated into the DBTL cycle to predict (Learn) and recommend (Design) variants of a bacterial RBS with the goal of optimising associated protein expression level in the specific genetic context (i.e. with specified bases upstream and downstream of the investigated RBS sequences). 
Two types of ML algorithms are applied in the DBTL cycle. 
For the Learn phase, the goal is to train a predictor that will predict the protein expression level by learning from the logged data. 
We used a Bayesian, non-parametric regression algorithm, called Gaussian Process Regression (GPR) \cite{Rasmussen2004} in our pipeline. 
GPR has been shown to  perform well with a small amount of data in biological prediction tasks \cite{Rasmussen2004, desautels2014parallelizing}. 
For the Design phase, the goal is to design a policy to recommend RBS sequences to query sequentially in batches so that we can identify the optimal RBS sequences within a given budget.
The policy uses the outputs of the GPR predictions to recommend useful batches for the Build and Test phases.
A variant of the Upper Confidence Bound multi-armed Bandit algorithms \cite{desautels2014parallelizing} was applied in our pipeline to guide our recommendations. 
The two types of algorithms cooperate with each other and provide powerful tools for DBTL cycle in genetic parts design \cite{srinivas2012information, desautels2014parallelizing}.
Using this approach, as summarized in Figure~\ref{fig: Flowchart}, we were able to increase expression of our target protein by up to 35\%, as compared to the very strong benchmark RBS. \\

@@

\hl{The exact position of the RBS in the sequence upstream of the protein coding sequence (CDS) can be hard to pinpoint, however most previous studies put it in the 20 bp directly preceding the coding sequence.} 
In our investigation, we are using an RBS which is known to have a very high TIR when expressing GFP and is present in the pBb series of plasmids \cite{Lee2011}.
This template RBS sequence is 20 base pairs (bp) long with the sequence TTTAAGA\textbf{AGGAGA}TATACAT, where the highlighted nucleotides constitute the core of the RBS.
Since this is the sequence against which new RBS sequences will be benchmarked,
we will refer to this sequence as the \textit{benchmark sequence} hereafter.
Additionally, we have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Figure \ref{fig:core_vs_noncore}).
This hypothesis has been built based on reported biases towards certain bases present in the core of the RBS but absent outside of it.
For example, according to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
Similarly, outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overall TIR of a given RBS. 
Focusing on the 6 to 8 bp core sequence is a common RBS design approach \cite{Jeschek2016}.
Hence, in our design, we focus on modification of the core at nucleotide positions -8 to -13 (relative to the start codon of the GFP; this is where the consensus Shine-Dalgarno AGGAGG sequence is usually found in wild type \textit{E. coli}) of the RBS and we keep other positions the same as the benchmark sequence, i.e. TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T). 
The total experimental (variant) space is then $4^6$ = 4096.\\

@@

Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group shows the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups, suggesting that the algorithms have a better understanding of the experimental space given more data.
The decreased result in the 0.9th quantile compared to the maximum value for the Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for that round compared to others.
We see this effect in Figure \ref{fig: Swarmplot and Quantplot}C (with details shown in Supplementary \ref{subsec: Exploration-exploitation visualisation} and Figure  \ref{fig:exploitationvsexplorationdetail}), where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
Those with a high predicted mean are coloured blue and represent exploitation, those coloured red are with high predicted uncertainty and represent exploration.
\hl{The implication of the fact that an RBS sequence chosen at random will have low TIR
(as shown by UNI and PPM), is that most of the exploration will result in low TIRs.
Our results confirm that RBSs with high TIRs tend to come from exploitation of the design space,
whereas the exploration points give relatively low TIRs.
Note that the exploration is necessary to expand our knowledge to the unknown parts of the design space and in effect allow us to exploit it better.}\\

@@

\hl{To provide the predicted TIR with the confidence interval needed for the UCB algorithm in the Design phase, we have selected a regression algorithm, Gaussian Process Regression (GPR),  which can predict not only the means, but also uncertainty estimations for our data points. }
% GPR, as a non-parametric Bayesian regression algorithm, assumes the reward function is sampled from a Gaussian Process and has been shown with good performance with a small amount of data in biological prediction tasks \cite{Rasmussen2004, desautels2014parallelizing}. Additionally, GPR has been studied with bandits algorithms, such as UCB and its variants, with both theoretical and empirical good recommendation performance (i.e. sub-linear regret upper bound) \cite{srinivas2012information, desautels2014parallelizing}.
Figure \ref{fig: Scatterplot} \hl{shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round. As expected, the predictions in Round 0 were poor due to the use of approximate data. 
The predictions improved for the subsequent rounds, 
the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.}
% Figure \ref{fig:scatter abc1 TT.} shows the performance of predictions for individual rounds.
\\

\hl{The regression task in the DBTL cycle is more challenging than the large-scale data-based regression tasks.
In the early iterations, we have a limited number of data points and relatively high variability due to the measurement noise. 
However, since the predictions are only used for recommendations in the Design phase, instead of precise predictions of the mean for each RBS, we only need to provide a valid ranking for both the predicted mean TIR and uncertainties, and thus the ranking of the UCB scores.
This would ensure that for each round we would be testing the required designs, even if we are not exactly accurate in terms of their numerical characteristics.
Because of that, we provide the evaluation of the Learn phase in terms of a ranking-based Spearman correlation coefficient, which has been shown to be a more suitable evaluation metric when the prediction is used for recommendation tasks than coefficient of determination ($R^2$) or Pearson correlation coefficient}
\cite{Schober2018,  Kang2019}.\\

Additionally, the 
% obtained values of the coefficient of determination and Spearman correlation coefficient 
performance of the Learn phase 
is also influenced by the exploration-exploitation balance in the Design phase.
In each round, we select some data points for exploration of the areas in which we have few or none tested data points.
For those exploration points, since the predictor never learns their label distributions and patterns, there is no chance for the predictor to provide accurate predictions of TIR values,
% which means that when tested, they have a high chance of having a experimentally-obtained mean different from what was predicted, 
thus hurting the prediction performance in the short term.
However, this is useful information for future predictions as it allows us to understand the whole underlying space instead of focusing on local sub-optimal data points.
\hl{In other words, we are intentionally sacrificing the accuracy of our predictions in each round to improve them in the future rounds}. 
The effect of exploration of the space is the ability to find high TIR RBSs even with relatively low prediction performance.\\

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_0.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_1.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_2.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/scatter_abc1_FF_3.pdf}
    \end{subfigure}
    \caption{\textbf{Performance of the prediction algorithm (no kernel normalisation)}. The scatter plots A-D show the performance of our prediction algorithm calculated after each round.
    Note that the TIR values are normalised according to the standardisation described in section \ref{sec: method data pre-procesing}, which is different from the TIR ratio reported in Figure \ref{fig: Swarmplot and Quantplot}.
    X-axis and y-axis are the true, tested TIR and the predicted TIR respectively.
    In (A)(B)(C), we show at round $t = 1,2,3$ respectively, we train our predictor based on the previously obtained data (round 0 to $t-1$) and show the predictions on both the training data (orange) and recommendations suggested by the Design phase (i.e. test data, blue). 
    In D), we train our predictor on randomly chosen 80\% of all available data, and test on the remaining 20\% data. 
    The Spearman correlation coefficient (with corresponding p-value; calculated using test data only) are provided in each plot's title on test data only.
    The p-value here is for the null hypothesis stating that two sets of data are uncorrelated.
    }
    \label{fig: Scatterplot}
\end{figure}

@@

\hl{Finally, while our strong sequences showed some affinity for the anti-sense sequence of the ribosome known to bind to RBS, they did not show any obvious secondary structures that could explain their TIRs} (see Figure \ref{fig:structures}).
This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion, based on the previously reported literature \cite{Saito2020,EspahBorujeni2016}, that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently \hl{optimise a small genetic part, in this case maximising the TIR of a bacterial RBS}. 
We have used Gaussian Process regression to predict the TIR function to be optimised and an Upper Confidence Bound multi-armed Bandit algorithm to recommend sequences to be tested.
To represent RBSs and capture the similarities between them, we chose to use the Weighted Degree Kernel with Shift method, which fits well into our prediction method.
In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow and testing only 450 RBS variants, we have designed and experimentally-validated RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs \hl{in this genetic context} by up to 34\%.
Furthermore, we have generated an extensive library of diverse RBSs that can be used \hl{as a basis} for future studies.\\

In our work we have focused on maximising the TIR of the RBS, by gradually moving our emphasis from exploration to exploitation as we progressed through the design rounds.
\hl{While maximisation could be the appropriate function for optimising RBSs or other small parts}, it may have to be modified in more complex cases, especially involving multi-gene metabolic engineering since maximising expression of genes would most probably result in excessive metabolic burden.
This change could be achieved within the Bayesian optimisation framework by changing the objective of the algorithm to maximising the range of the black-box function, or identifying the part of function above a threshold \mbox{\cite{gotovos2013active}}, or combining different goals into a multi-objective method, for example \mbox{\cite{shu2020new}}.\\
% \hl{
% While we focus our optimisation framework on the RBS optimisation task, our framework has good potential to be generalised to multi-gene pathway design, as reviewed by \mbox{\textcite{LAWSON2021}} and proposed by \mbox{\textcite{hamedirad2019towards}}. For example, the optimisation goal can be adjusted to address combinatorial optimisation for multi-gene and RBS scenarios; since this would be a large-scale data task, the current Gaussian Process Regression (GPR) prediction model could be updated to a deep GPR approach and the current Bandit algorithm could be optimised towards querying large design space to reduce computational complexity \mbox{\cite{shahriari2015taking}}.  
% }

@@

There are still open questions that need to be addressed for applying machine learning in synthetic biology.
Firstly, we would like to understand how we can extract more biologically-important information from the decisions made by our algorithms.
We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the results obtained.
Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty measurements? 
Thirdly, the generalisability of the method is unknown.
We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
\hl{Similarly, the re-usability of the obtained data set is currently unknown. 
The TIR of an RBS is dependent on its genetic context, but our Hamming distance and TIR impact of nucleotides outside of the core (Figures} \ref{fig:Library characteristics}B and \ref{fig:core_vs_noncore} \hl{respectively) studies indicate that as long as the changes in the genetic context are small the obtained data set could serve as a basis for similar design efforts.
For example, the data could be used to teach the algorithm for the first round of new designs, which in turn could decrease the number of rounds required to obtain the required characteristics.}
Finally, the optimal exploration-exploitation balance between rounds and samples is still to be determined.
In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.\\
